---
title: "Covid-19 India Forecast"
author: "Baan Bapat"
date: "26 May 2020"
output:
  html_document:
    theme: cerulean
    code_folding: hide
    self_contained: false
---
<style>
h1,h2{page-break-before: always}
</style>

##### Assumptions

* Models are based on severe lockdown period. 
* Unfortunately, all indications are for relaxation of lockdown in the coming week and the assumption is already breaking down.
* Effect(s) of migration unknown.

##### Credits

* Data source: api.covid19india.org



```{r, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, comment=NA, 
                      fig.align='center', fig.height=5, fig.width=10) 
                      # dev.args=list(pointsize=10), out.height="0.5\\paperheight", out.width="0.7\\linewidth")
options(stringsAsFactors=FALSE, knitr.kable.NA = '', scipen=9)
library(data.table)
library(jsonlite)
library(ggplot2)
library(plotly)
library(gridExtra)

#' Get json data from covid19india.org
df = read_json("https://api.covid19india.org/data.json")
sf0 = read_json("https://api.covid19india.org/states_daily.json")

```


```{r}
#' Some data points used
#' rolling windows
rolldays = 3
#' Lockdown dates
lockdowns = as.Date(c('2020-03-25', '2020-04-14', '2020-05-03', 
                      '2020-05-17', '2020-05-31'))
#' Start date for model training data
startdt = lockdowns[3]
#' How many days ahead forecast
ahead = 21
```

```{r}
cts <- function(){
  #' extract cases_time_series
  cases_time_series = rbindlist(df[['cases_time_series']])
  cases_time_series[, `:=`(
    date = as.Date(date, '%d %B'),
    dailyconfirmed = as.integer(dailyconfirmed),
    dailydeceased = as.integer(dailydeceased),
    dailyrecovered = as.integer(dailyrecovered),
    totalconfirmed = as.integer(totalconfirmed),
    totalrecovered = as.integer(totalrecovered),
    totaldeceased = as.integer(totaldeceased))]
  cases_time_series[, `:=`(active = totalconfirmed - totaldeceased - totalrecovered,
                           fatality = totaldeceased / totalconfirmed * 100)]

  setcolorder(cases_time_series, 'date')
  #' Daily case rate for total & recovered
  cases_time_series[, dailyrate := dailyconfirmed/totalconfirmed]
  cases_time_series[, recovrate := dailyrecovered/totalrecovered]
  #' Rolling rates of the same two for smoothing
  cases_time_series[, rollrate := frollmean(dailyrate, n=rolldays)]
  cases_time_series[, recovrr := frollmean(recovrate, n=rolldays)]
  #' Doubling rate -- only for confirmed and recovered
  cases_time_series[, doublingrate := log(2)/log(1 + rollrate)]
  cases_time_series[, recovdr := log(2)/log(1 + recovrate)]
  cases_time_series = cases_time_series[date>=lockdowns[1],]
  return(cases_time_series)
}


  

  #' Overall & State-wise data today's status
  #' active, confirmed, deaths, deltaconfirmed, deltadeaths, deltarecovered,
  #' lastupdatedtime, recovered, state
  statewise = rbindlist(df[['statewise']])
  # statewise[,1:9]
  #' Tested
  tested = rbindlist(df[['tested']])

  #' State case data
get_sf <- function(){
  sf = rbindlist(sf0[[1]])
  sf = melt(sf, id.vars=c('date', 'status'), value.name='cases',
            variable.name='state')
  sf[,  `:=`(cases = as.integer(cases),
             date = as.Date(date, '%d-%b-%y'))]
  sf[is.na(cases), cases:=0]
  sf = dcast(sf, date + state ~ status, value.var='cases', fill=0)
  setorder(sf, 'state', 'date')
  sf[, `:=`(totalconfirmed = cumsum(Confirmed),
            totaldeceased = cumsum(Deceased),
            totalrecovered = cumsum(Recovered)), by='state']
  sf[, active := totalconfirmed - totalrecovered - totaldeceased]
  sf[, `:=` (dailyrate  = Confirmed/pmax(totalconfirmed, 1),
             recovrate  = Recovered/pmax(totalrecovered, 1),
             fatality = totaldeceased / totalconfirmed * 100)]
  sf[, `:=`(rollrate = frollmean(dailyrate, n=3),
            recovrollrate = frollmean(recovrate, n=3)), by='state']
  sf[, `:=`(doublingrate = log(2)/log(1 + dailyrate),
            rolldr = log(2)/log(1 + rollrate),
            recovdr = log(2)/log(1 + recovrollrate))]
  sf = sf[date>=lockdowns[1] + 6,]

  # Average doubling rate over first week  May
  dr = sf[date >= Sys.Date() - 7,
          .(doublingrate =  log(2)/log(1 + mean(dailyrate)),
            rolldr =  log(2)/log(1 + mean(rollrate)),
            recovrolldr =  log(2)/log(1 + mean(recovrollrate))),
          by='state']
  #' States with highest current active cases
  topca = sf[date==max(sf$date) - 1, .(state, active, totalconfirmed)]
  dr = dr[topca, on='state']
  dr = dr[doublingrate < 20  & active >= 100,]
  setorder(dr, rolldr)
  return(list(sf = sf, dr = dr))
}


cases_time_series = cts()
dtupto = max(cases_time_series$date)
sfd = get_sf()
sf = sfd$sf
dr = sfd$dr

```


## All India case count


```{r}
g = ggplot(melt(cases_time_series[, .(date, totalconfirmed, totaldeceased,
                                      totalrecovered, active)], id.vars='date',
                value.name='cases', variable.name='series'),
           aes(x=date, y=cases, col=series)) +
  geom_line() + 
  geom_vline(xintercept=lockdowns, col='gray50') +
  scale_x_date(breaks = scales::breaks_pretty(30)) +
  scale_y_continuous(trans="log10") +
  theme(axis.text.x = element_text(angle = 90)) + xlab('Date') +
  ylab('Total confirmed cases')
ggplotly(g)
```

* Plotted on a logarithmic scale
* Gray lines indicate the lockdown dates
* Shows a fairly straight line trend in total confirmed, recovered & deceased from the third lockdown (`r format(lockdowns[3], '%d %b %Y')`).

## Fatality

```{r}

ggplot(cases_time_series, aes(x=fatality)) + 
  geom_histogram(bins=40, col='gray50') + ylab('Count of days')
```

```{r}
print(summary(cases_time_series$fatality))
```

```{r}
#' Case fatality rate
cfr = median(cases_time_series$fatality)
```

* Defining fataility on a cummulative daily basis as $\frac{\text{total deceased}}{\text{total confirmed}}\times 100$.
* We may consider median fataility  as case fatailty rate (cfr = `r round(cfr,2)`).

## Predicting the next three weeks

* We fit a trend model to log of total confirmed, recovered & deceased.
* We predict `r ahead` days ahead using all of May data from `r format(lockdowns[3], '%d %b %Y')` up to `r format(dtupto, '%d %b %Y')`
* Active cases are predicted as difference of confirmed and the other two.
* Note a marked jump and change of slope in recovered and active cases 16 May onwards. This seems to also make the model fit not so good for these two series. 
    - Recovery appears to be over predicted and active case count under predicted 



```{r}
#' Fitting regression model to the log of May data using
pi_jun <- function(trn){
  trn[, t := 1:.N]
  pred = data.table(t = 1:(nrow(trn)+ahead),
                    date = seq(startdt, by=1,
                               length.out=(nrow(trn)+ahead)))
  getPred <- function(x){
    fml = as.formula(paste0("log(", x, ") ~ t"))
    m1 = lm(fml, trn)
    print(paste("log", x))
    print(summary(m1))
    pd = predict(m1, newdata=pred[,"t"], 
                 interval="prediction")
    pd = as.data.table(apply(pd, 2, function(y) 
      round(exp(y))))
    pd$series=paste0('pred_', substr(x, 6, 9))
    cbind(pred, pd)
  }
  pred = rbind(getPred('totalconfirmed'),
               getPred('totalrecovered'),
               getPred('totaldeceased'))
  pred[, t := NULL]
  setnames(pred, 'fit', 'cases')
  actv = copy(pred[series=='pred_conf', ])
  actv[, `:=`(cases = cases - 
                pred[series=='pred_reco',]$cases -
                pred[series=='pred_dece',]$cases,
              lwr = lwr - pred[series=='pred_reco',]$lwr -
                pred[series=='pred_dece',]$lwr,
              upr = upr - pred[series=='pred_reco',]$upr -
                pred[series=='pred_dece',]$upr,
              series = 'pred_activ')]
  trn[, t := NULL]
  trn = melt(trn, id.vars='date', value.name='cases', 
                    variable.name='series')
  trn[, `:=`(lwr = NA, upr = NA)]
  setcolorder(trn, names(pred))
  pred = rbind(trn, pred, actv)
  pred[, series := factor(series, levels = c(levels(pred$series)[5:8], 
                                            levels(pred$series)[1:4]))]
}
jpi = pi_jun(trn = cases_time_series[date>=startdt, 
                                       .(date, totalconfirmed, totalrecovered,
                                         totaldeceased, active)])
```

## Forecast summary (`r ahead` days from `r format(dtupto, '%d %b %Y')`)


```{r}
cols = c('#c2a5cf', '#a6dba0', '#f4a582', '#92c5de',
         '#7b3294', '#008837', '#ca0020', '#0571b0')
legend = jpi[date==max(jpi$date), .(date, series, cases)]
setorder(legend, 'series')
legend[, series := c('confirmed', 'recovered', 'deceased', 'active')]
gi = ggplot(jpi, aes(x=date, y=cases, col=series)) + geom_line() +
  geom_vline(xintercept=lockdowns[3:4], col='gray50') +
  geom_ribbon(aes(ymin=lwr, ymax=upr, fill=series), alpha=0.2) +
  geom_text(mapping=aes(label=series, x=date, y=cases, hjust=1, vjust=0),
            data=legend, col='gray50', nudge_x=-3, nudge_y=1.5) +
  scale_color_manual(values=cols) +
  scale_x_date(breaks = scales::breaks_pretty(30)) +
  theme(axis.text.x = element_text(angle = 90), legend.position="none") + xlab('Date') +
  ylab('Count of cases')
ggplotly(gi)
```

* By May-end we expect:
    - Active cases will be less than recovered
    - One lakh active cases & Two lakh total cases
* Three lakh total cases by 7 June
* Active cases: 1.2 lakh by 9 June


## Maharashtra case count


```{r}
g = ggplot(melt(sf[state=='mh', .(date, totalconfirmed, totaldeceased,
                                      totalrecovered, active)], id.vars='date',
                value.name='cases', variable.name='series'),
           aes(x=date, y=cases, col=series)) +
  geom_line() + 
  geom_vline(xintercept=lockdowns[-1], col='gray50') +
  scale_x_date(breaks = scales::breaks_pretty(30)) +
  scale_y_continuous(trans="log10") +
  theme(axis.text.x = element_text(angle = 90)) + xlab('Date') +
  ylab('Total confirmed cases')
ggplotly(g)
```

## Maharashtra Fatality

```{r}

ggplot(sf[state=='mh'], aes(x=fatality)) + 
  geom_histogram(bins=40, col='gray50') + ylab('Count of days')
```

```{r}
print(summary(sf[state=='mh',]$fatality))
```



```{r}
#' Case fatality rate
cfrm = median(sf[state=='mh',]$fatality)
```

* Maharashtra case fatailty rate (cfrm = `r round(cfrm,2)`).

## Predicting the next three weeks

* We predict `r ahead` days ahead using all of May data from `r format(lockdowns[3] + 6, '%d %b %Y')` up to `r format(dtupto, '%d %b %Y')`


```{r}
junm = pi_jun(trn = sf[state=='mh' & date>=startdt, 
                       .(date, totalconfirmed, totalrecovered,
                         totaldeceased, active)])

```

## Maharashtra forecast summary (`r ahead` days from `r format(dtupto, '%d %b %Y')`)

```{r}
legend = junm[date==max(junm$date), .(date, series, cases)]
setorder(legend, 'series')
legend[, series := c('confirmed', 'recovered', 'deceased', 'active')]
gm = ggplot(junm, aes(x=date, y=cases, col=series)) + geom_line() +
  geom_vline(xintercept=lockdowns[3:4], col='gray50') +
  geom_ribbon(aes(ymin=lwr, ymax=upr, fill=series), alpha=0.2) +
  geom_text(mapping=aes(label=series, x=date, y=cases, hjust=1, vjust=0),
            data=legend, col='gray50', nudge_x=-3, nudge_y=1.5) +
  scale_color_manual(values=cols) +
  scale_x_date(breaks = scales::breaks_pretty(30)) +
  theme(axis.text.x = element_text(angle = 90), legend.position="none") + xlab('Date') +
  ylab('Count of cases')
ggplotly(gm)
```

* Total cases forecast: 1 lakh by 4 Jun, 1.5 lakh by 11 Jun & 2 lakh by 15 Jun
* Recovered case count forecast to exceed active case count by 11 Jun
* Active case count forecast: 50k by 1 Jun, 75k by 11 Jun

## Rest of India (excluding Maharashtra)



```{r}
ri = copy(cases_time_series[date >= lockdowns[1] + 6, 
                            .(date, totalconfirmed, totaldeceased,
                              totalrecovered, active)])
ri = ri[sf[state=='mh', ], `:=`(totalconfirmed = totalconfirmed - i.totalconfirmed,
                                totaldeceased = totaldeceased - i.totaldeceased,
                                totalrecovered = totalrecovered - i.totalrecovered,
                                active = active - i.active), on='date']
ri[, fatality := totaldeceased / totalconfirmed * 100]
```


```{r}

g = ggplot(melt(ri[, .(date, totalconfirmed, totaldeceased,
                                      totalrecovered, active)], id.vars='date',
                value.name='cases', variable.name='series'),
           aes(x=date, y=cases, col=series)) +
  geom_line() + 
  geom_vline(xintercept=lockdowns[-1], col='gray50') +
  scale_x_date(breaks = scales::breaks_pretty(30)) +
  scale_y_continuous(trans="log10") +
  theme(axis.text.x = element_text(angle = 90)) + xlab('Date') +
  ylab('Total confirmed cases')
ggplotly(g)
```

## Rest of India Fatality

```{r}

ggplot(ri, aes(x=fatality)) + 
  geom_histogram(bins=40, col='gray50') + ylab('Count of days')
```

```{r}
print(summary(ri$fatality))
```



```{r}
#' Case fatality rate
cfri = median(sf[state=='mh',]$fatality)
```


* Rest of India case fatailty rate (cfr = `r round(cfri,2)`).

## Predicting the next three weeks

* We predict `r ahead` days ahead using all of May data from `r format(lockdowns[3] + 6, '%d %b %Y')` up to `r format(dtupto, '%d %b %Y')`


```{r}
jri = pi_jun(trn = ri[date>=startdt, .(date, totalconfirmed, totalrecovered,
                                          totaldeceased, active)])
```

## Rest of India forecast summary (`r ahead` days from `r format(dtupto, '%d %b %Y')`)

```{r}
legend = jri[date==max(jri$date), .(date, series, cases)]
setorder(legend, 'series')
legend[, series := c('confirmed', 'recovered', 'deceased', 'active')]
gri = ggplot(jri, aes(x=date, y=cases, col=series)) + geom_line() +
  geom_vline(xintercept=lockdowns[3:4], col='gray50') +
  geom_ribbon(aes(ymin=lwr, ymax=upr, fill=series), alpha=0.2) +
  geom_text(mapping=aes(label=series, x=date, y=cases, hjust=1, vjust=0),
            data=legend, col='gray50', nudge_x=-3, nudge_y=1.5) +
  scale_color_manual(values=cols) +
  scale_x_date(breaks = scales::breaks_pretty(30)) +
  theme(axis.text.x = element_text(angle = 90), legend.position="none") + xlab('Date') +
  ylab('Count of cases')
ggplotly(gri)
```

* Total cases forecast: 1.5 lakh by 3 Jun & 2 lakh by 9 Jun
* Recovered case count has just crossed active case count
* Active case count forecast to peak at about 51k on 6 Jun, and start dropping thereafter(!)


```{r}
jp = rbind(jpi[, grp:='India'],
           junm[, grp:='Maharashtra'],
           jri[, grp:='Ind ex. Maha'])
jp[, grp:=factor(grp, levels=c('India', 'Maharashtra', 'Ind ex. Maha'))]
```


## Forecast Covid-19 trends (May 25-Jun15 2020)

```{r}
legend = jpi[date==max(jpi$date), .(date, series, cases)]
setorder(legend, 'series')
legend[, series := c('confirmed', 'recovered', 'deceased', 'active')]
legend[, grp:=factor('India', levels=c('India', 'Maharashtra', 'Ind ex. Maha'))]
g = ggplot(jp, aes(x=date, y=cases, col=series)) + geom_line() +
  geom_vline(xintercept=lockdowns[3:4], col='gray50') +
  geom_ribbon(aes(ymin=lwr, ymax=upr, fill=series), alpha=0.2) +
  geom_text(mapping=aes(label=series, x=date, y=cases, hjust=0, vjust=0),
            data=legend, col='gray50', nudge_x=-5, nudge_y=1.5) +
  scale_color_manual(values=cols) +
  scale_x_date(breaks = scales::breaks_pretty(6)) +
  facet_wrap(~grp) +
  theme(axis.text.x = element_text(angle = 90), legend.position="none") +
  xlab('') + ylab('')
ggplotly(g)
```

```{r}
smy = data.frame("Feature" = c('Recovered cases more than Active cases', 'Active cases as on 15 Jun', 'Confirmed cases as on 15 Jun',
                               'When do Active cases peak'),
                 "India" = c("1 Jun", '1,27,325', '4,57,461', 'Not by 15 Jun'),
                 "Maharashtra" = c("11 Jun", '85,789', '1,96,027', 'Not by 15 Jun'),
                 "India Ex. Maha" = c("22 May", '43,016', '2,68,380', '6 Jun (?)'))
knitr::kable(smy)
```

* Forecast based on 1 May - 24 May 2020 data (source: covid19india.org) -- complete lockdown period.
* Model used: trend fit to log of cases (confirmed, deceased or recovered) -- Lockdown assumption may not be valid in forecast period.
